---
title: "Alpha Diversity Basics"
output:
  pdf_document: default
  html_notebook: default
---

## Introduction

Now that we have some data to work with the we can get to work. In this exercise we will learn how to quantify alpha diversity (species diversity) of microbial community data.In all the diversity analysis scripts we will extensively use the vegan package. 
>The vegan package provides tools for descriptive community ecology. It has most basic functions of diversity analysis, community ordination and dissimilarity analysis. Most of its multivariate tools can be used for other data types as well (Taken from the vegan help page).
  

*** 
## Analysis

### Initialize project and generate some random data

To start we will import the libraries needed and use random distribution generators to make a neg binomial distributed matrix this will be a reasonable-ish approximation of microbiome data. For simplicity, a small handmade data set will also be used. 

```{r data_import}
library(vegan)

#Make small data set

df <- data.frame(patient_1 = c(5,5,5,5,5,5),
                 patient_2 = c(0,0,0,0,0,1000),
                 patient_3 = c(0,50,5,100,10,1),
                 patient_4 = c(0,0,5,5,5,0),
                 patient_5 = c(100,100,100,100,100,100))

rownames(df) <- c("species_a",
                  "species_b",
                  "species_c",
                  "species_d",
                  "species_e",
                  "species_f")


# Use random distribution generators to make a neg binomial distributed 
# matrix this will be a reasonable-ish approximation of microbiome data

set.seed(731)
big_df <- matrix(rnbinom(10000, size = .2, prob = .01),
                 nrow = 100,
                 ncol = 100)

colnames(big_df) <- paste("patient", 1:100,sep = "_" )
rownames(big_df) <- paste("species", 1:100,sep = "_" )

```

***

### Normalization

Because varying sequencing depth and unequal input concentration is unavoidable in microbiome science, normalization needs to be applied to account for this. Two common methods are rarefaction where a data set is randomly resampled to a common sequencing depth, and relative abundance where a proportional microbial abundances are calculated for each sample independently.  

```{r normalization}


#Rarefaction is necessary to conduct alpha diversity analysis
rarecurve(x = df,
          step = 10, # step is how often you sample
          )

#set sample size
sample = min(colSums(big_df)) /10 # you want enough sample points to make smooth

rarecurve(x = big_df,
          step = floor(sample),
          label = FALSE

)

abline(v = 100)
abline(v = 500)
abline(v = 1100) # looks like a few sample

#remove all samples < 1100
big_df.rare <- big_df[,which(colSums(big_df) > 1099)]

#remove all species == 0
big_df.rare <- big_df.rare[which(rowSums(big_df) > 0),]

# rrarefy, note the second "r", requires sampling the rows not columns
# the t() function transposes data (i.e., rows become columns)

big_df.rare <- rrarefy(t(big_df.rare), sample = 1100)

#check to make sure all
all(colSums(big_df.rare) > 0)


big_df.rel <- big_df[,which(colSums(big_df) > 1099)]

big_df.rel <- sweep(x = big_df.rel,
                    MARGIN = 2,
                    STATS = colSums(big_df.rel),
                    FUN = '/'
                    )
big_df.rel <- t(big_df.rel)

#How different are relative abundance and rarefied data tables?

cor.test(matrix(big_df.rare),matrix(big_df.rel))
cor.test(vegdist(big_df.rare),vegdist(big_df.rel))

#df will not be rarefied

```

***

### Estimate Richness 

Species richness is a simple alpha diversity metric that quantifies the total number of species, OTUs, ASV, phylotypes, etc. within a given sample. It is (usually) strongly influenced by normalization method with raw, and relative abundance data having increased richness.

```{r richness}
# The function in R::vegan to determine richness is specnumber

specnumber(x = df,
          MARGIN = 2 #by columns
          )

#raw
big_df.richness <- specnumber(x = big_df,
                    MARGIN = 2 #by columns
                    )

#rarefied
big_df_rare.richness <- specnumber(x = big_df.rare,
                    MARGIN = 1 #by columns
                    )

#how are the values distributed for normalized and raw data
boxplot(big_df.richness, big_df_rare.richness,
        names = c("raw", "rarefied"))
hist(big_df.richness)
hist(big_df_rare.richness) #rarefied data frame
```

***

### Shannon Diversity

Shannon diversity is a measure of how "random" a community is. As number of species and evenness increases so does Shannon diversity.  

```{r shannon_diversity }


diversity(df,
          index = "shannon",
          MARGIN = 2
          )

big_df.shannon <- diversity(big_df,
          index = "shannon",
          MARGIN = 2
)

big_df_rare.shannon <- diversity(big_df.rare,
                            index = "shannon",
                            MARGIN = 1
)

#how are the values distributed for normalized and raw data

boxplot(big_df.shannon, big_df_rare.shannon,
        names = c("raw", "rarefied"))
hist(big_df.shannon)
hist(big_df_rare.shannon)
```

***
### Simpson Diversity Index
The Simpson index is another popular alpha-diversity metric. It is accessible through the diversity function. 

```{r simpson_diversity}

diversity(df,
          index = "simpson",
          MARGIN = 2
)

big_df.simpson <- diversity(big_df,
                            index = "simpson",
                            MARGIN = 2
)

big_df_rare.simpson <- diversity(big_df.rare,
                                 index = "simpson",
                                 MARGIN = 1
)

#how are the values distributed for normalized and raw data
boxplot(big_df.simpson, big_df_rare.simpson,
        names = c("raw", "rarefied"))
hist(big_df.simpson)
hist(big_df_rare.simpson)
```

*** 
#### Relative abundance correlations 

Different metrics yield different results with varying degrees of correlation which depend on the underlying data structure.  

```{r correlations}

#Linear relationships cannot be assumed so we will use nonparametic

#Richness vs simpson
plot(big_df_rare.richness, big_df_rare.simpson)
cor.test(big_df_rare.richness, big_df_rare.simpson, method = "spearman")

#Richness vs shannon

plot(big_df_rare.richness, big_df_rare.shannon)
cor.test(big_df_rare.richness, big_df_rare.shannon, method = "spearman")

#simpson vs shannon

plot(big_df_rare.simpson, big_df_rare.shannon)
cor.test(big_df_rare.simpson, big_df_rare.shannon, method = "spearman")
```



