---
title: "Beta Diversity Basics"
output:
  pdf_document: default
  html_notebook: default
---


## Introduction

One of the first analysis conducted in a microbiome study is a beta diversity analysis. This end point of this analysis is usually some form of ordination. Such ordinations have come to be expected in microbiome papers and in this tutorial we will go over some basics on how to produce ordinations using popular methods. 

*** 
## Analysis

### Initialize project and generate some random data

To start we will import the libraries needed and use random distribution generators to make a neg binomial distributed matrix this will be a reasonable-ish approximation of microbiome data. By splitting this into two deferentially parameterized distributions we will be able to visualize differences later.

```{r}
library(vegan)   # for diversity analyses  
library(ggplot2) # for visuals
library(ape)     # for PCoA

#make reproducible data set
set.seed(731)
big_df1 <- matrix(rnbinom(10000, size = .05, prob = .001),
                 nrow = 100,
                 ncol = 100)

big_df2 <- matrix(rnbinom(10000, size = .05, prob = .001),
                  nrow = 100,
                  ncol = 100)

colnames(big_df1) <- paste("patient", 1:100,sep = "_" )
rownames(big_df1) <- paste("species", 1:100,sep = "_" )

colnames(big_df2) <- paste("patient", 101:200,sep = "_" )
rownames(big_df2) <- paste("species", 51:150,sep = "_" )


big_df <- merge(big_df1,big_df2,by  = 0, all =T )
big_df[is.na(big_df)] <- 0
rownames(big_df) <- big_df$Row.names
big_df$Row.names <- NULL
dim(big_df)

```
***
###  Normalize data 

Normalization is important as discussed before. Here we will just use a simple rarefaction approach.   

```{r}


big_df.rare <- big_df[,which(colSums(big_df) >= min(colSums(big_df)))]

#remove all species == 0
big_df.rare <- big_df.rare[which(rowSums(big_df) > 0),]

# rrarefy, note the second "r", requires sampling the rows not colums
# the t() function transposes data (i.e., rows become columns)

big_df.rare <- rrarefy(t(big_df.rare), sample = min(colSums(big_df)))

#check to make sure all are there should be 150 x 200 
all(colSums(big_df.rare) > 0)

```
***
### Calculating beta diversity 

There are many different methods for calculating diversity. We are going to concentrate on using two that are widely used in microbiome analyses: jaccard and bray-curtis. Since the most come implementation of these metrics is non-binary (i.e., abundance weighted) we will focus on these, however, you should be aware that binary versions of these algorithms are implemented by vegan. The Prior to trying this yourself you should read more about the diversity metrics available in vegan (e.g., ?vegdist).

```{r}

#Bray Curtis
big_df_rare.bray <- vegdist(x = big_df.rare,
                            method = "bray", #bray-curtis is the method
                            binary = FALSE, # weighted
                            upper = FALSE  # only compure 
                            )

big_df_rare.jacc <- vegdist(x = big_df.rare,
                            method = "jaccard",
                            binary = FALSE,
                            upper = FALSE)

# for these data the distances are similar, but for other data it can 
# make a big difference
cor.test(big_df_rare.bray, big_df_rare.jacc)
plot(big_df_rare.bray, big_df_rare.jacc,xlim = c(0,1), ylim=c(0,1))

```

***

### Principle Components Analysis (PCA)

Next we are going to see if the species distributions of these patient samples stratify our artificially generated data sets. For PCA, a diversity matrix is not necessary. 

```{r}

big_df.prcomp <- prcomp(big_df.rare, center = T, scale =T)

#to determine how much variance is explained lets look at the summary
summary(big_df.prcomp)$importance[,c(1:5)]

big_df_prcomp.df <- as.data.frame(big_df.prcomp$x[,1:5])
big_df_prcomp.df$group <- rep(c(1,2), times = c(100,100))


big_df_prcomp.plot <- ggplot(big_df_prcomp.df,
                             aes(x = PC1,
                                 y = PC2,
                                 color = factor(group))
                             )

big_df_prcomp.plot+ geom_point()

```

***

### Principle Coordinate Analysis (PCoA)

PCoA is similar to PCA in that it attempts to find the axes of variation that explain the largest portion of variation in the whole data set. Unlike PCA, PCoA requires a diversity matrix (aka distance matrix generated with vegdist) and the selection of diversity metric will impact the results. 

```{r}

# PCoA wants a matrix of dist object. We will provide both distance 
# objects separately, starting with bray

big_df_bray.pcoa <- pcoa(big_df_rare.bray)

#With PCoA the we don't get percent variation but an analogous value of
# brokenstick can be accessed
big_df_bray.pcoa$values[c(1:5),]

big_df_bray_pcoa.df <- as.data.frame(big_df_bray.pcoa$vectors[,1:5])
big_df_bray_pcoa.df$group <- rep(c(1,2), times = c(100,100))

big_df_bray_pcoa.plot <- ggplot(big_df_bray_pcoa.df,
                             aes(x = Axis.1,
                                 y = Axis.2,
                                 color = factor(group))
)

big_df_bray_pcoa.plot + 
  geom_point()

#now jaccard

big_df_jacc.pcoa <- pcoa(big_df_rare.jacc)

big_df_jacc_pcoa.df <- as.data.frame(big_df_jacc.pcoa$vectors[,1:5])
big_df_jacc_pcoa.df$group <- rep(c(1,2), times = c(100,100))

big_df_jacc_pcoa.plot <- ggplot(big_df_jacc_pcoa.df,
                                aes(x = Axis.1,
                                    y = Axis.2,
                                    color = factor(group))
)

big_df_jacc_pcoa.plot + 
  geom_point()
```

***

### Non-metric multidimensional scaling (NMDS)

NMDS is similar to both PCA and PCoA in that it also will try to find the axes of variation that explain the largest portion of variation. However, unlike PCA and PCoA, NMDS is not based on eignevector based decomposition of data frames and NMDS is rank based. Another key difference is that it is more difficult to determine how much variation is explained  by each axis. 

```{r}

# Now we will try the same approach but with NMDS. For NMDS we have to 
# decide on a number of dimensions. The default is 2, but usually that 
# will yield a very high stress (bad). To attempt to minimize stress
# we will try a few values of k (dimensions) and find the one the results
# in lowest stressour target is around or below .1

(metaMDS(big_df.rare,distance = "bray", 
         k = 2,
         trace = FALSE))$stress # too high to be useful
(metaMDS(big_df.rare,distance = "bray",
         k = 5,
         trace = FALSE))$stress # better, but still high
(metaMDS(big_df.rare,
         distance = "bray",
         k = 10,
         trace = FALSE))$stress # better, but still high
(metaMDS(big_df.rare,
         distance = "bray",
         k = 12,
         trace = FALSE))$stress # good


big_df_rare_bray.NMDS <- metaMDS(big_df.rare,
                                 distance = "bray",
                                 k = 12,
                                 trace = FALSE)

# Unlike PCA and PCoA, no % of variance associated with each axis in NMDS

big_df_rare_bray_nmds.df <- as.data.frame(big_df_rare_bray.NMDS$points)
big_df_rare_bray_nmds.df$group <- rep(c(1,2), times = c(100,100))

big_df_rare_bray_nmds.plot <- ggplot(big_df_rare_bray_nmds.df,
                                aes(x = MDS1,
                                    y = MDS2,
                                    color = factor(group))
)

big_df_rare_bray_nmds.plot + 
  geom_point()

#now jaccard
big_df_rare_jacc.NMDS <- metaMDS(big_df.rare,
                                 distance = "jaccard", 
                                 k = 12,
                                 trace = FALSE)
big_df_rare_jacc_nmds.df <- as.data.frame(big_df_rare_jacc.NMDS$points)

big_df_rare_jacc_nmds.df$group <- rep(c(1,2), times = c(100,100))

big_df_rare_jacc_nmds.plot <- ggplot(big_df_rare_jacc_nmds.df,
                                     aes(x = MDS1,
                                         y = MDS2,
                                         color = factor(group))
)

big_df_rare_jacc_nmds.plot +
  geom_point()

```

